---
title: "Final Project"
format: html
editor: source
---

## Method

```{r}
# TODO: add libraries
library(tidyverse)
library(plotly)

# load data
covid <- read_csv("../data/covid_cases_deaths.csv")
```

### Data Preparation

To prepare and wrangle the data for this analysis, we integrated multiple datasets, cleaned the data, and structured it for consistency and analysis. The key steps are detailed below:

#### 1. **Population Data Wrangling**

-   We processed population data for U.S. states from two datasets: population estimates from 2020 to 2023 and 2024 projections. These datasets were merged into a single dataset.

-   The states with missing population data for 2024 was filled using its corresponding 2023 population.

-   State names were converted to standardized abbreviations for consistency, and non-U.S. states (e.g., territories) were excluded.

-   The population data was pivoted into a long format, where each row represents a state-year combination with the corresponding population value.

#### 2. **COVID-19 Death Data Wrangling**

-   The COVID-19 death dataset was filtered to include only weekly data for consistency.

-   Relevant columns, including weekly start and end dates, state, COVID-19 deaths, total deaths, and percentage of expected deaths, were renamed and standardized.

-   Dates were converted into a standard date format, and year and week numbers were extracted.

-   State names were standardized to abbreviations, and data was filtered to include only states present in the population dataset.

-   The cleaned dataset includes weekly COVID-19 death counts and related information for each state.

#### 3. **Combining Death and Population Data**

-   The cleaned COVID-19 death data was merged with the population dataset by state and year to ensure each record included population data, allowing the calculation of rates (e.g., death rates per 100,000 people).

#### 4. **COVID-19 Cases Data Wrangling**

-   The COVID-19 cases dataset was cleaned and standardized. Weekly case numbers, end dates, and state abbreviations were extracted and converted into the same structure as the death data.

-   Only states present in the population dataset were included.

-   The processed dataset contains weekly case counts for each state.

#### 5. **Final Dataset Integration**

-   The cleaned COVID-19 death, population, and case datasets were integrated into a single dataset by merging them on state, year, and week.

-   Population data was forward and backward filled to ensure completeness for all records.

-   Additional columns, such as epidemiological week (`epiweek`) and year, were calculated for consistency with standard reporting practices.

#### 6. **Final Dataset Output**

-   The final dataset contains the following variables: state, date, epidemiological week, COVID-19 deaths, total deaths, population, and cases. This dataset provides a comprehensive, time-resolved view of COVID-19 impacts across U.S. states.

-   The processed dataset was exported as a CSV file for subsequent analysis.

### 1) Divide the pandemic period into waves

Divide the pandemic period, January 2020 to December 2024 into waves. Justify your choice with data visualization.

We first computed the following:

-   COVID-19 death rate per 100,000 people

-   Total death rate per 100,000 people (for the following questions)

-   Case fatality rate (%) (for the following questions)

```{r}
covid_overall <- covid|>
  group_by(year, week)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            total_death = sum(total_death, na.rm = TRUE),
            population = sum(population), 
            cases_total = sum(cases, na.rm = TRUE),
            date = first(date),
            .groups = 'drop')|> 
  mutate(covid_death_rate_100k = (covid_death / population) * 1e5,
         total_death_rate_100k = (total_death / population) * 1e5,
         case_fatality_rate = (covid_death/cases_total)*100)

base_plot <- covid_overall|> 
  ggplot(aes(x = date,))+
  labs(x = "Date")
```

### 1.1 COVID-19 Death Rate Trend (2020-2024)

We then made two plots to observe the trends of COVID-19 death rate from 2020 to 2024 and COVID-19 cases from 2020 to 2023.

```{r}
y_map_death <- aes(y = covid_death_rate_100k)

death_rate_plot <- base_plot + 
  geom_point(mapping = y_map_death) + 
  geom_line(mapping = y_map_death, color = "red") +
  ylab("COVID Death Rate Per 100K")

ggplotly(death_rate_plot)
```

### 1.2 COVID-19 Case Trend (2020-2023)

```{r}
y_map_cases <- aes(y = cases_total)

cases_plot <- base_plot + 
  geom_point(mapping = y_map_cases) + 
  geom_line(mapping = y_map_cases, color = "red") +
  ylab("Total COVID Cases")

ggplotly(cases_plot)
```

### 1.3 Define Periods

After analyzing the trends in COVID-19 death rates and case counts, we decided to divide the pandemic period into three waves based on the number of COVID-19 cases. This approach was chosen because it better reflects the overall spread of the virus. For instance, the virus might have been spreading widely among younger populations, who are less likely to experience severe outcomes or fatalities. By focusing on case counts rather than death tolls, we aim to capture the dynamics of viral transmission during the pandemic.

-   Period 1: 2020-01-11 to 2021-11-06

-   Period 2: 2021-11-06 to 2022-04-09

-   Period 3: 2022-04-09 to 2024-05-13

```{r}
covid_period_1 <- covid|>
  filter(date >= "2020-01-11" & date <= "2021-11-06")
  
covid_period_2 <- covid|>
  filter(date >= "2021-11-06" & date <= "2022-04-09")

covid_period_3 <- covid|>
  filter(date >= "2022-04-09" & date <= "2024-05-13")
```

### 2) COVID 19 Death Rates by State

For each period, compute the death rates by state. Describe which states did better or worse during the different periods.

### 2.1 Period 1: 2020-01-11 to 2021-11-06

##### 2.1.1 State-Level COVID-19 Death Rate

```{r}
p2.1.1 <-covid_period_1|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "State-Level COVID-19 Death Rate (2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.1.1)
```

##### 2.1.2 State-Level Cumulative COVID-19 Death Rates

```{r}
p2.1.2 <- covid_period_1 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "State-Level Cumulative COVID-19 Death Rates (2020-01-11 to 2021-11-06)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.1.2)
```

#### Observation:

During Period 1, Tennessee (TN) had the greatest COVID-19 death rate when Vermont (VT) had the lowest.

### **2.2 Period 2: 2021-11-06 to 2022-04-09**

##### 2.2.1 State-Level COVID-19 Death Rate

```{r}
p2.2.1 <-covid_period_2|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "State-Level COVID-19 Death Rate (2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.2.1)
```

##### 2.2.2 State-Level Cumulative COVID-19 Death Rates

```{r}
p2.2.2 <- covid_period_2 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "State-Level Cumulative COVID-19 Death Rates (2021-11-06 to 2022-04-09)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.2.2)
```

#### Observation:

During Period 2, West Virginia (WV) had the greatest COVID-19 death rate when Hawaii (HI) had the lowest.

### 2.3 Period 3: 2022-04-09 to 2024-05-13

##### 2.3.1 State-Level COVID-19 Death Rate

```{r}
p2.3.1 <-covid_period_3|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "State-Level COVID-19 Death Rate (2022-04-09 to 2024-05-13)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.3.1)
```

##### 2.3.2 State-Level Cumulative COVID-19 Death Rates

```{r}
p2.3.2 <- covid_period_3 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "State-Level Cumulative COVID-19 Death Rates (2022-04-09 to 2024-05-13)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.3.2)
```

#### Observation:

During Period 3, Kentucky (KY) had the greatest COVID-19 death rate when Alaska (AK) had the lowest.

### 3) COVID 19 Trend

Describe if COVID-19 became less or more virulent across the different periods.

### 3.1 Period 1: 2020-01-11 to 2021-11-06

##### 3.1.1 National COVID-19 Case Fatality Rate

```{r}
p3.1.1 <- covid_period_1|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "National COVID-19 Case Fatality Rate (2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.1.1)
```

#### Observation:

During Period 1, the COVID-19 fatality rate exhibited a sharp increase from February 29, 2020, to March 7, 2020, peaking at 20.79%. Following this peak, the fatality rate declined significantly to 2.59% and eventually to 0.77% towards the end of this period.

The trend during this period showed that COVID-19 became less virulent over time. However, the sharp rise and fall during Period 1 indicate that early in the pandemic, the virus was particularly lethal, possibly due to a lack of understanding of effective treatments and overwhelmed healthcare systems.

Notably, according to **Plot 3.1.2**, Florida (FL) recorded the highest COVID-19 fatality rate (77.27%) among all states.

##### 3.1.2 State-Level COVID-19 Case Fatality Rates

```{r}
p3.1.2 <- covid_period_1|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "State-Level COVID-19 Case Fatality Rates (2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.1.2)
```

### **3.2 Period 2: 2021-11-06 to 2022-04-09**

##### 3.2.1 National COVID-19 Case Fatality Rate

```{r}
p3.2.1 <- covid_period_2|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "National COVID-19 Case Fatality Rate (2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.2.1)
```

#### Observation:

During Period 2, the COVID-19 fatality rate exhibited an unstable trend, initially decreasing, followed by an increase, and then another decrease. The highest fatality rate during this period was 1.68%, and the lowest was 0.38%, both significantly lower than the corresponding highest and lowest rates in Period 1.

The fatality rate experienced two major decreases and one major increase during this period. The first decrease occurred between November 6, 2021, and January 15, 2022, while the second decrease was observed between March 12, 2022, and April 9, 2022. The major increase took place between January 15, 2022, and March 12, 2022.

Notably, according to **Plot 3.2.2,** Kansas (KS) recorded the highest COVID-19 fatality rate (7.02%) among all states.

##### 3.2.2 State-Level COVID-19 Case Fatality Rates

```{r}
p3.2.2 <- covid_period_2|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "State-Level COVID-19 Case Fatality Rates (2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.2.2)
```

### 3.3 Period 3: 2022-04-09 to 2024-05-13

##### 3.3.1 National COVID-19 Case Fatality Rate

```{r}
p3.3.1 <- covid_period_3|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "National COVID-19 Case Fatality Rate (2022-04-09 to 2024-05-13)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.3.1)
```

#### Observation:

During Period 3, the COVID-19 fatality rate initially exhibited a sharp decrease and followed by a gradual increase. The decline occurred between April 9, 2022 and May 21, 2022, and the increase then followed. The highest fatality rate during this period was 1.07%, and the lowest was 0.25%, both significantly lower than the corresponding highest and lowest rates in Period 1 and slighly lower than the ones in Period 3.

Notably, according to **Plot 3.1.2**, Kentucky (KY) recorded the highest COVID-19 fatality rate (8.58%) among all states.

##### 3.3.2 State-Level COVID-19 Case Fatality Rates

```{r}
p3.3.2 <- covid_period_3|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "State-Level COVID-19 Case Fatality Rates (2022-04-09 to 2024-05-13)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.3.2)
```

```{r}
#### 5) Weekly Excess Mortality by State

Estimate excess mortality for each week for each state. Do COVID-19 deaths explain the excess mortality?

week_2024 <- covid_death_clean|>
  filter(year == 2024)|>
  select(mmwr_week)|>
  na.omit()

dat <- covid_death_clean|>
  left_join(cases_population_state, by = c("state", "year"))|>
  rename(week = "mmwr_week")|>
  mutate(mortality = total_death/population*1e5,
         day = difftime(start_date, min(start_date), units = "day"),
         week = as.factor(week))|>
  filter(week %in% week_2024$mmwr_week)|>
  select(-cases_total)|>
  group_by(state)|>
  fill(covid_death, total_death, population, .direction = "down")

fit <- dat|> 
  filter(year == "2024") #2024 because it has the lowest COVID 19 Mortality Rate

mod <- lm(mortality ~ week+day, fit)

pred <- predict(mod, newdata = dat, se.fit = TRUE)
```

```{r}
dat <- dat|>
  mutate(expected = pred$fit*population,
         excess = total_death - expected,
         se_expected = pred$se.fit*population,
         sigma_outcome = sd(mod$residuals)*population,
         se_excess = sqrt(sigma_outcome^2 + se_expected^2))
```

```{r}
covid_death_excess <- covid_death_clean|>
  left_join(expected_death_year, by = c("state", "year"))|>
  mutate(excess_mortality = expected_death - total_death)

covid_death_excess|>
  select(year, mmwr_week, state, excess_mortality, covid_death, total_death)
```

#### 6) Excess Mortality by State

Repeat 2 but for excess mortality instead of COVID-19 deaths.

```{r}
# Compute average excess mortality
avg_excess <- covid_death_excess |> 
  group_by(state) |> 
  mutate(avg_excess = mean(excess_mortality, na.rm = TRUE)) |> 
  filter(mmwr_week == max(mmwr_week, na.rm = TRUE)) |> 
  ungroup() |> 
  arrange(desc(excess_mortality))
  

#2020-2022 excess mortality
ggplot(avg_excess, 
       aes(x = avg_excess, y = reorder(state, avg_excess))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Excess Mortality by State from 2020-2022",
    x = "Average Excess Mortality",
    y = "State"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
weekly_excess_2022 <- covid_death|>
  filter(Year == 2022)|>
  group_by(state) |> 
  select(mmwr_week, state, covid_deaths, total_death) |> 
  filter(complete.cases(mmwr_week, state, covid_deaths, total_death)) |> 
  mutate(excess_mortality = total_death-covid_deaths) 
  
#2022 excess mortality by week
ggplot(weekly_excess_2022, 
       aes(x = mmwr_week, y = excess_mortality, color = state)) +
  geom_line() +
  labs(
    title = "Excess Mortality by Week in 2022",
    x = "Week",
    y = "Excess Mortality"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
