---
title: "Final Project"
format: html
editor: source
---

```{r}
# TODO: add libraries
library(tidyverse)
library(plotly)

# load data
covid <- read_csv("../data/covid_cases_deaths.csv")
```


#### 2) Divide the pandemic period into waves

Divide the pandemic period, January 2020 to December 2024 into waves. Justify your choice with data visualization.

#### COVID-19 Death Rate per 100,000 People Over Time (2020-2024)

##### Plot 1: COVID-19 Death Rate per 100,000 People by State (2020-2024)

```{r}
covid_overall <- covid|>
  group_by(year, week)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            total_death = sum(total_death, na.rm = TRUE),
            population = sum(population), 
            cases_total = sum(cases, na.rm = TRUE),
            date = first(date),
            .groups = 'drop')|> 
  mutate(covid_death_rate_100k = (covid_death / population) * 1e5,
         total_death_rate_100k = (total_death / population) * 1e5,
         case_fatality_rate = (covid_death/cases_total)*100)

base_plot <- covid_overall |> ggplot(aes(
  x = date,
)) + labs(
  x = "Date"
)

y_map_death <- aes(y = covid_death_rate_100k)
death_rate_plot <- base_plot + 
  geom_point(mapping = y_map_death) + 
  geom_line(mapping = y_map_death) +
  ylab("COVID Death Rate Per 100K")

ggplotly(death_rate_plot)

y_map_cases <- aes(y = cases_total)
cases_plot <- base_plot + 
  geom_point(mapping = y_map_cases) + 
  geom_line(mapping = y_map_cases) +
  ylab("Total COVID Cases")

ggplotly(cases_plot)
```

Notes on rationale:

* we plotted both cases and deaths over time

* we decided to just focus on the cases to determine the waves 

  * We made this choice because we feel it better reflects the state of the virus. For example, the virus could be spreading a lot among young people but not killing any of them since they are young. We wanted to focus on how it was spreading rather than the death toll. 

* Periods: start of pandemic - 2021-11-06; 2021-11-06 - 2022-04-09; 2022-04-09 - 2024-05-13 (when no more case data)

##### Plot 2: COVID-19 Death Rate per 100,000 People in the US (2020-2024)

```{r}
p2 <- covid_overall|>
  ggplot(aes(x=year, y=covid_death_rate_100k))+
  geom_line(color = "red")+
  geom_point(color="blue")+
  labs(
    title = "COVID-19 Death Rate per 100,000 People in the US (2020-2024)",
    x = "Year",
    y = "Death Rate (per 100,000)")

ggplotly(p2)
```

#### Observation:

-   First wave: 2020

-   Second wave: Late 2020 to early 2021

#### COVID-19 Cases Over Time (2020-2023)

##### Plot 3: COVID-19 Cases by State (2020-2023)

```{r}
p3 <- covid_state|>
  filter(year!="2024")|>
  mutate(year=as.numeric(year))|>
  ggplot(aes(x=year, y=cases_total, color=state))+
  geom_point()+
  geom_line()+
  labs(title = "COVID-19 Cases by State (2020-2023)",
       x="Year",
       y="Cases")

ggplotly(p3)
```

##### Plot 4: COVID-19 Cases in the US (2020-2023)

```{r}
p4 <- covid_overall|>
  filter(year != "2024")|>
  ggplot(aes(x=year, y=cases_total))+
  geom_point(color="red")+
  geom_line(color="blue")+
  labs(title = "COVID-19 Cases in the US (2020-2023)",
       x="Year",
       y="Cases")

ggplotly(p4)
```

#### Observation:

-   Highest peak: 2022 (wave 2021-2022)

-   Second peak: 2021 (wave 2020-2021)

##### Plot 5: COVID-19 Case Rate vs COVID-19 Death Rate (2020-2023)

```{r}
#Compare COVID-19 cases and deaths on the same plot

p5 <- covid_overall|>
  filter(year != "2024")|> # case data is not availbe in 2024
  mutate(covid_case_rate_100k =(cases_total/population)*1e5)|>
  ggplot(aes(x=year))+
  geom_line(aes(y=covid_case_rate_100k, color="Case Rate"))+
  geom_line(aes(y=covid_death_rate_100k, color="Death Rate"))+
  labs(title = "COVID-19 Case Rate vs COVID-19 Death Rate (2020-2023)",
       x="Year",
       y="Rate per 100,000",
       color= "Metric")+
  scale_color_manual(values = c("Case Rate" = "navy",
                                "Death Rate" = "red"),
                     name="")

ggplotly(p5)
```

#### 3) COVID 19 Death Rates by State

For each period, compute the death rates by state. Describe which states did better or worse during the different periods.

##### Plot 6: COVID-19 Death Rates by State (2021-2022)

```{r}
p6 <- covid_state|>
  filter(year>=2021 & year<=2022)|>
  ggplot(aes(x =covid_death_rate_100k, y = reorder(state, covid_death_rate_100k)))+
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "COVID-19 Death Rates by State (2021-2022)",
       x = "Death Rate per 100,000",
       y = "State")

ggplotly(p6)
```

-   Observations for Death Rate per 100,000 (2021-2022):

    -   Highest: West Virginia (WV)

    -   Lowest: Vermont (VT)

##### Plot 7: COVID-19 Death Rates by State (2021)

```{r}
p7 <- covid_state|>
  filter(year=="2021")|>
  ggplot(aes(x =covid_death_rate_100k, y = reorder(state, covid_death_rate_100k)))+
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "COVID-19 Death Rates by State (2021)",
       x = "Death Rate per 100,000",
       y = "State")

ggplotly(p7)
```

-   Observations for Death Rate per 100,000 (2021):

    -   Highest: West Virginia (WV)

    -   Lowest: Vermont (VT)

##### Plot 8: COVID-19 Death Rates by State (2022)

```{r}
p8 <- covid_state|>
  filter(year=="2022")|>
  ggplot(aes(x =covid_death_rate_100k, y = reorder(state, covid_death_rate_100k)))+
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "COVID-19 Death Rates by State (2022)",
       x = "Death Rate per 100,000",
       y = "State")

ggplotly(p8)
```

-   Observations for Death Rate per 100,000 (2022):

    -   Highest: West Virginia (WV)

    -   Lowest: Alaska (AK)

#### 4) COVID 19 Trend

Describe if COVID-19 became less or more virulent across the different periods.

**COVID-19 Case Fatality Rate by State and Overall (2020-2023)**

##### Plot 9: COVID-19 Case Fatality Rate by States (2020-2023)

##### Plot 10: COVID-19 Case Fatality Rate in the US (2020-2023)

```{r}
p9 <- covid_state|>
  filter(year!="2024")|> # case data is not availbe in 2024
  ggplot(aes(x=year, y=case_fatality_rate, colour = state))+
  geom_point()+
  geom_line()+
  labs(title="COVID-19 Case Fatality Rate by States (2020-2023)",
       x="Year",
       y="Case Fatality Rate (%)")
  
ggplotly(p9)

p10 <- covid_overall|>
  filter(year!="2024")|> # case data is not availbe in 2024
  ggplot(aes(x=year, y=case_fatality_rate))+
  geom_point(color="blue")+
  geom_line(color="red")+
  labs(title="COVID-19 Case Fatality Rate in the US (2020-2023)",
       x="Year",
       y="Case Fatality Rate (%)")

ggplotly(p10)
```

-   Observations for COVID-19 Case Fatality Rate in the US (2020-2023):

    -   CFR experienced a sharp decline from 2020 to 2021 and continued to decrease at a slower pace from 2021 to 2022.

    -   However, there was an increase in the CFR from 2022 to 2023.

##### Plot 11: COVID-19 Death Rates by State (2020-2021)

##### Plot 12: COVID-19 Case Fatality Rate by States (2021-2022)

```{r}
p11 <- covid_state|>
  filter(year>="2020" & year<="2021")|>
  ggplot(aes(x=year, y=case_fatality_rate, colour = state))+
  geom_point()+
  geom_line()+
  labs(title="COVID-19 Case Fatality Rate by States (2020-2021)",
       x="Year",
       y="Case Fatality Rate (%)")

p12 <- covid_state|>
  filter(year>="2021" & year<="2022")|>
  ggplot(aes(x=year, y=case_fatality_rate, colour = state))+
  geom_point()+
  geom_line()+
  labs(title="COVID-19 Case Fatality Rate by States (2021-2022)",
       x="Year",
       y="Case Fatality Rate (%)")

ggplotly(p11)
ggplotly(p12)
```

-   Observations for COVID-19 Case Fatality Rate by States (2020-2022):

    -   Vermont exhibited the highest CFR in early 2020. However, from 2021 to 2022, there was a dramatic decrease in its CFR, ultimately making it the state with the lowest CFR.

-   Observations for COVID-19 Case Fatality Rate by States (2021-2022):

    -   Nevada exihbited the highest CFR in early 2021 while Vermont exhibited the lowest.

##### Plot 13: COVID-19 Case Fatality Rate in the US (2020-2021)

##### Plot 14: COVID-19 Case Fatality Rate in the US (2021-2022)

```{r}
p13 <- covid_overall|>
  filter(year>="2020" & year<="2021")|>
  ggplot(aes(x=year, y=case_fatality_rate))+
  geom_point(color="blue")+
  geom_line(color="red")+
  scale_x_continuous(breaks = 2020:2021, 
                     labels = as.character(2020:2021))+
  labs(title="COVID-19 Case Fatality Rate in the US (2020-2021)",
       x="Year",
       y="Case Fatality Rate (%)")

p14 <- covid_overall|>
  filter(year>="2021" & year<="2022")|>
  ggplot(aes(x=year, y=case_fatality_rate))+
  geom_point(color="blue")+
  geom_line(color="red")+
  scale_x_continuous(breaks = 2021:2022, 
                     labels = as.character(2021:2022))+
  labs(title="COVID-19 Case Fatality Rate in the US (2021-2022)",
       x="Year",
       y="Case Fatality Rate (%)")

ggplotly(p13)
ggplotly(p14)
```

-   Observations for COVID-19 Case Fatality Rate in the US (2020-2021) and (2021-2022):

    -   For both periods, CFR exhibited a decline.

#### 5) Weekly Excess Mortality by State

Estimate excess mortality for each week for each state. Do COVID-19 deaths explain the excess mortality?

```{r}
week_2024 <- covid_death_clean|>
  filter(year == 2024)|>
  select(mmwr_week)|>
  na.omit()

dat <- covid_death_clean|>
  left_join(cases_population_state, by = c("state", "year"))|>
  rename(week = "mmwr_week")|>
  mutate(mortality = total_death/population*1e5,
         day = difftime(start_date, min(start_date), units = "day"),
         week = as.factor(week))|>
  filter(week %in% week_2024$mmwr_week)|>
  select(-cases_total)|>
  group_by(state)|>
  fill(covid_death, total_death, population, .direction = "down")

fit <- dat|> 
  filter(year == "2024") #2024 because it has the lowest COVID 19 Mortality Rate

mod <- lm(mortality ~ week+day, fit)

pred <- predict(mod, newdata = dat, se.fit = TRUE)
```

```{r}
dat <- dat|>
  mutate(expected = pred$fit*population,
         excess = total_death - expected,
         se_expected = pred$se.fit*population,
         sigma_outcome = sd(mod$residuals)*population,
         se_excess = sqrt(sigma_outcome^2 + se_expected^2))
```

```{r}
covid_death_excess <- covid_death_clean|>
  left_join(expected_death_year, by = c("state", "year"))|>
  mutate(excess_mortality = expected_death - total_death)

covid_death_excess|>
  select(year, mmwr_week, state, excess_mortality, covid_death, total_death)
```

#### 6) Excess Mortality by State

Repeat 2 but for excess mortality instead of COVID-19 deaths.

```{r}
# Compute average excess mortality
avg_excess <- covid_death_excess |> 
  group_by(state) |> 
  mutate(avg_excess = mean(excess_mortality, na.rm = TRUE)) |> 
  filter(mmwr_week == max(mmwr_week, na.rm = TRUE)) |> 
  ungroup() |> 
  arrange(desc(excess_mortality))
  

#2020-2022 excess mortality
ggplot(avg_excess, 
       aes(x = avg_excess, y = reorder(state, avg_excess))) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Excess Mortality by State from 2020-2022",
    x = "Average Excess Mortality",
    y = "State"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
weekly_excess_2022 <- covid_death|>
  filter(Year == 2022)|>
  group_by(state) |> 
  select(mmwr_week, state, covid_deaths, total_death) |> 
  filter(complete.cases(mmwr_week, state, covid_deaths, total_death)) |> 
  mutate(excess_mortality = total_death-covid_deaths) 
  
#2022 excess mortality by week
ggplot(weekly_excess_2022, 
       aes(x = mmwr_week, y = excess_mortality, color = state)) +
  geom_line() +
  labs(
    title = "Excess Mortality by Week in 2022",
    x = "Week",
    y = "Excess Mortality"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
