---
title: "Supplementary Methods"
format: html
editor: visual
---

## **Supplementary Methods**

```{r}
#| output: false

library(tidyverse)
library(plotly)

# load data
covid <- read_csv("../data/covid_cases_deaths.csv")
```

### 1) COVID-19 Pandemic Waves

#### Figure 1. **COVID-19 death rate from 2020 to 2024**

```{r}
base_plot <- covid|>
  group_by(year, week)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE), #Total COVID-19 cases by week 
            cases_total = sum(cases, na.rm = TRUE),
            population = sum(population),
            date = first(date),
            .groups = 'drop')|> 
  mutate(covid_death_rate_100k = (covid_death / population) * 1e5)|> #COVID-19 death rate per 100,000 people by week
  ggplot(aes(x = date,))+
  labs(x = "Date")
```

```{r}
#COVID-19 Death Rate Trend (2020-2024)
y_map_death <- aes(y = covid_death_rate_100k)

p1.1 <- base_plot + 
  geom_point(mapping = y_map_death) + 
  geom_line(mapping = y_map_death, color = "red") +
  labs(title = "Figure 1. COVID-19 death rate from 2020 to 2024",
       y = "COVID Death Rate Per 100K")

ggplotly(p1.1)
```

#### Figure 2. **COVID-19 cases from 2020 to 2023**

```{r}
#COVID-19 Case Trend (2020-2023)
y_map_cases <- aes(y = cases_total)

p1.2 <- base_plot + 
  geom_point(mapping = y_map_cases) + 
  geom_line(mapping = y_map_cases, color = "red") +
  labs(title = "Figure 2. COVID-19 cases from 2020 to 2023",
       y = "Total COVID Cases")

ggplotly(p1.2)
```

### 2) COVID 19 Death Rates by State Comparison

```{r}
covid_period_1 <- covid|>
  filter(date >= "2020-01-11" & date < "2021-11-06")
  
covid_period_2 <- covid|>
  filter(date >= "2021-11-06" & date < "2022-04-09")

covid_period_3 <- covid|>
  filter(date >= "2022-04-09" & date <= "2023-05-13")
```

### 2.1 Period 1

#### Figure 3. State-Level COVID-19 Death Rate (2020-01-11 to 2021-11-06)

```{r}
# 2.1.1 State-Level COVID-19 Death Rate
p2.1.1 <-covid_period_1|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 3. State-Level COVID-19 Death Rate (2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.1.1)
```

#### Figure 4. State-Level Cumulative COVID-19 Death Rates

```{r}
#2.1.2 State-Level Cumulative COVID-19 Death Rates
p2.1.2 <- covid_period_1 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "Figure 4. State-Level Cumulative COVID-19 Death Rates\n(2020-01-11 to 2021-11-06)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.1.2)
```

### 2.2 Period 2

#### Figure 5. State-Level COVID-19 Death Rate (2021-11-06 to 2022-04-09)

```{r}
#2.2.1 State-Level COVID-19 Death Rate
p2.2.1 <-covid_period_2|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 5. State-Level COVID-19 Death Rate (2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.2.1)
```

#### Figure 6. State-Level Cumulative COVID-19 Death Rates (2021-11-06 to 2022-04-09)

```{r}
#2.2.2 State-Level Cumulative COVID-19 Death Rates
p2.2.2 <- covid_period_2 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "Figure 6. State-Level Cumulative COVID-19 Death Rates\n(2021-11-06 to 2022-04-09)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.2.2)
```

### 2.3 Period 3

#### Figure 7. State-Level COVID-19 Death Rate (2022-04-09 to 2023-05-13)

```{r}
#2.3.1 State-Level COVID-19 Death Rate
p2.3.1 <-covid_period_3|>
  mutate(covid_death_rate_100k = (covid_death / population)*1e5)|>
  ggplot(aes(x=date, y=covid_death_rate_100k, color= state))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 7. State-Level COVID-19 Death Rate\n(2022-04-09 to 2023-05-13)",
       x = "Date",
       y = "Death Rate (per 100k People)")

ggplotly(p2.3.1)
```

#### Figure 8. State-Level Cumulative COVID-19 Death Rates (2022-04-09 to 2023-05-13)

```{r}
#2.3.2 State-Level Cumulative COVID-19 Death Rates
p2.3.2 <- covid_period_3 |> 
  group_by(state) |>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            population = first(population),            
            death_rate_per_100k = (covid_death /population)*1e5,
            .groups = 'drop')|>
  ggplot(aes(x=reorder(state, death_rate_per_100k), y=death_rate_per_100k))+
  geom_bar(stat = 'identity', fill = 'lightblue')+
  coord_flip()+
  labs(title = "Figure 8. State-Level Cumulative COVID-19 Death Rates\n(2022-04-09 to 2023-05-13)",
       x = "State",
       y = "Death Rate (per 100k people)")

ggplotly(p2.3.2)
```

### 3) COVID 19 Virulence Trend

### 3.1 Period 1

#### Figure 9. National COVID-19 Case Fatality Rate (2020-01-11 to 2021-11-06)

```{r}
#3.1.1 National COVID-19 Case Fatality Rate
p3.1.1 <- covid_period_1|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 9. National COVID-19 Case Fatality Rate (2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.1.1)
```

#### Figure 10. State-Level COVID-19 Case Fatality Rates (2020-01-11 to 2021-11-06)

```{r}
#3.1.2 State-Level COVID-19 Case Fatality Rates
p3.1.2 <- covid_period_1|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "Figure 10. State-Level COVID-19 Case Fatality Rates\n(2020-01-11 to 2021-11-06)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.1.2)
```

### 3.2 Period 2

#### Figure 11. National COVID-19 Case Fatality Rate (2021-11-06 to 2022-04-09)

```{r}
#3.2.1 National COVID-19 Case Fatality Rate
p3.2.1 <- covid_period_2|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 11. National COVID-19 Case Fatality Rate\n(2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.2.1)
```

#### Figure 12. State-Level COVID-19 Case Fatality Rates (2021-11-06 to 2022-04-09)

```{r}
#3.2.2 State-Level COVID-19 Case Fatality Rates
p3.2.2 <- covid_period_2|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "Figure 12. State-Level COVID-19 Case Fatality Rates\n(2021-11-06 to 2022-04-09)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.2.2)
```

### 3.3 Period 3

#### Figure 13. National COVID-19 Case Fatality Rate (2022-04-09 to 2023-05-13)

```{r}
#3.3.1 National COVID-19 Case Fatality Rate
p3.3.1 <- covid_period_3|>
  group_by(date)|>
  summarise(covid_death = sum(covid_death, na.rm = TRUE),
            cases_total = sum(cases, na.rm = TRUE),
            .groups = 'drop')|>
  mutate(case_fatality_rate = (covid_death/cases_total)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate))+
  geom_point()+
  geom_line()+
  labs(title = "Figure 13. National COVID-19 Case Fatality Rate\n(2022-04-09 to 2023-05-13)",
       x = "Date",
       y = "Case Fatality Rate (%)")

ggplotly(p3.3.1)

```

#### Figure 14. State-Level COVID-19 Case Fatality Rates (2022-04-09 to 2023-05-13)

```{r}
#3.3.2 State-Level COVID-19 Case Fatality Rates
p3.3.2 <- covid_period_3|>
  mutate(case_fatality_rate = (covid_death/cases)*100)|>
  ggplot(aes(x=date, y=case_fatality_rate, color=state))+
  geom_point()+
  geom_line()+
 labs(title = "Figure 14. State-Level COVID-19 Case Fatality Rates\n(2022-04-09 to 2023-05-13)",
       x = "Date",
       y = "Case Fatality Rate (%)")
  
ggplotly(p3.3.2)
```

### 4) Weekly Excess Mortality by State

#### Figure 15. Overall Excess Mortality and COVID Death Rates in U.S. Throughout Pandemic

```{r}
# compute excess mortality
covid_exp_deaths <- covid |> 
  # compute expected deaths
  mutate(prop_exp_death = percentage_expected_death / 100) |> 
  mutate(exp_death = total_death / prop_exp_death) |> 
  # compute excess mortality
  # need to use rates since mortality will depend on population
  mutate(
    total_death_rate = total_death / population * 1e5,
    exp_death_rate = exp_death / population * 1e5
  ) |> 
  mutate(exc_mort_rate = total_death_rate - exp_death_rate)

# plot excess mortality
# for simplicity of viewing, will collapse to total excess mortality over all US
p4 <- covid_exp_deaths |> 
  select(state, date, covid_death, total_death, exp_death, population) |> 
  group_by(date) |> 
  summarize(
    covid_death = sum(covid_death, na.rm=T),
    total_death = sum(total_death, na.rm=T),
    exp_death = sum(exp_death, na.rm=T),
    population = sum(population, na.rm=T) # since we are collapsing across states, sum population of each state to get total US pop
  ) |> 
  ungroup() |> 
  mutate(
    total_death_rate = total_death / population * 1e5,
    exp_death_rate = exp_death / population * 1e5,
    covid_death_rate = covid_death / population * 1e5
  ) |> 
  mutate(exc_mort_rate = total_death_rate - exp_death_rate) |> 
  select(date, covid_death_rate, exc_mort_rate) |> 
  pivot_longer(cols = c("covid_death_rate", "exc_mort_rate"), names_to = "death_type", values_to = "num_deaths") |> 
  mutate(death_type = case_when(
    death_type == "covid_death_rate" ~ "COVID Death Rate",
    death_type == "exc_mort_rate" ~ "Excess Mortality Rate"
  )) |> 
  ggplot(aes(
    x = date,
    y = num_deaths,
    color = death_type
  )) + geom_point() + geom_line() +
  labs(
    x = "Date",
    y = "Death Rate (per 100,000)",
    color = "Death Rate Type",
    title = "Figure 15. Overall Excess Mortality and COVID Death Rates in U.S.\nThroughout Pandemic"
  )
 
ggplotly(p4) 
```

### 5) Excess Mortality by State Comparison

#### Figure 16. Excess Mortality Rate By State During Period 1 (2020-01-11 to 2021-11-06)

```{r}
covid_pers <- list(
  per1 = list(
    start = make_date(2020, 1, 11),
    end = make_date(2021, 11, 6)
  ),
  per2 = list(
    start = make_date(2021, 11, 6),
    end = make_date(2022, 4, 9)
  ),
  per3 = list(
    start = make_date(2022, 4, 9),
    end = make_date(2023, 5, 13)
  )
)

# aggregate across all dates
exc_mortality_period <- function(per_start_date, per_end_date) {
  output <- covid_exp_deaths |> 
    filter((date >= per_start_date) & (date < per_end_date)) |> 
    group_by(state) |> 
    summarize(
      covid_death = sum(covid_death, na.rm=T),
      total_death = sum(total_death, na.rm=T),
      exp_death = sum(exp_death, na.rm=T),
      population = first(population)
    ) |> ungroup() |> 
    mutate(
      total_death_rate = total_death / population * 1e5,
      exp_death_rate = exp_death / population * 1e5,
      covid_death_rate = exp_death / population * 1e5
    ) |> 
    mutate(exc_mort_rate = total_death_rate - exp_death_rate)
  
  return(output)
}

#Period 1
exc_mort_per_1 <- exc_mortality_period(covid_pers$per1$start, covid_pers$per1$end)
p5.1 <- exc_mort_per_1 |> mutate(state = reorder(state, exc_mort_rate)) |> 
  ggplot(aes(y = state, x = exc_mort_rate)) +
  geom_col() + labs(
    x = "Excess Mortality Rate (per 100,000 people)",
    y = "State",
    title = "Figure 16. Excess Mortality Rate By State During Period 1\n(2020-01-11 to 2021-11-06)"
  )
ggplotly(p5.1)
```

#### Figure 17. Excess Mortality Rate By State During Period 2 (2021-11-06 to 2022-04-09)

```{r}
#Period 2
exc_mort_per_2 <- exc_mortality_period(covid_pers$per2$start, covid_pers$per2$end)
p5.2 <- exc_mort_per_2 |> mutate(state = reorder(state, exc_mort_rate)) |> 
  ggplot(aes(y = state, x = exc_mort_rate)) +
  geom_col() + labs(
    x = "Excess Mortality Rate (per 100,000 people)",
    y = "State",
    title = "Figure 17. Excess Mortality Rate By State During Period 2\n(2021-11-06 to 2022-04-09)"
  )
ggplotly(p5.2)
```

#### Figure 18. Excess Mortality Rate By State During Period 3 (2022-04-09 to 2023-05-13)

```{r}
#Period 3
exc_mort_per_3 <- exc_mortality_period(covid_pers$per3$start, covid_pers$per3$end)
p5.3 <- exc_mort_per_3 |> mutate(state = reorder(state, exc_mort_rate)) |> 
  ggplot(aes(y = state, x = exc_mort_rate)) +
  geom_col() + labs(
    x = "Excess Mortality Rate (per 100,000 people)",
    y = "State",
    title = "Figure 18. Excess Mortality Rate By State During Period 3\n(2022-04-09 to 2023-05-13)"
  )
ggplotly(p5.3)
```
